---
title: "COVID-19 Vaccination Rates"
author: "Duc Nguyen"
format: html
---

# Getting Started

## Import vaccination data

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")

head(vax)
```

> **Q1. What column details the total number of people fully vaccinated?**

```{r}
names(vax)
```

> **A1: persons_fully_vaccinated**

> **Q2. What column details the Zip code tabulation area?**

> **A2: zip_code_tabulation_area**

> **Q3. What is the earliest date in this dataset?**

```{r}
head(vax$as_of_date)
```

> **A3: 2021-01-05**

> **Q4. What is the latest date in this dataset?**

```{r}
tail(vax$as_of_date)
```

> **A4: 2022-11-15**

 Let’s call the 'skim()' function from the skimr package to get a quick overview of this dataset

```{r, message=FALSE}
skimr::skim(vax)
```

> **Q5. How many numeric columns are in this dataset?**

> **A5: 13**

> **Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?**

```{r}
na_persons_fully_vaccinated <- sum(is.na(vax$persons_fully_vaccinated))
na_persons_fully_vaccinated
```

> **A6: 15440**

> **Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?**

```{r}
round((na_persons_fully_vaccinated/nrow(vax))*100, 2)
```

> **A7: 8.93%**

> **Q8. [Optional]: Why might this data be missing?**

> **A8: I think that this data be missing might due to some people did not get the vaccine.**

## Working with dates

Install the "lubridate" package in R Console by using the 'install.packages("lubridate")' command.

```{r, message=FALSE}
library(lubridate)
```

What is today’s date

```{r}
today()
```

Convert our date data in the as_of_date column into a lubridate format (year-month-day)

```{r}
vax$as_of_date <- ymd(vax$as_of_date)
```

How many days have passed since the first vaccination reported in this dataset?

```{r}
today()
vax$as_of_date[1]
today() - vax$as_of_date[1]
```

Using the last and the first date value we can now determine how many days the dataset span?

```{r}
vax$as_of_date[nrow(vax)]
vax$as_of_date[1]
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

> **Q9. How many days have passed since the last update of the dataset?**

```{r}
today()
vax$as_of_date[nrow(vax)]
today() - vax$as_of_date[nrow(vax)]
```

> **A9: 7 days (as of today is 2022-11-22)**

>**Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?**

```{r}
length(unique(vax$as_of_date))
```

> **A10: 98 unique dates**

# Working with ZIP codes

Install the "zipcodeR" package in R Console to work with postal ZIP codes by using the 'install.packages("zipcodeR")' command.

```{r, warning=FALSE}
library(zipcodeR)
```

```{r}
geocode_zip('92037')
```

Calculate the distance between the centroids of any two ZIP codes in miles, e.g.

```{r}
zip_distance('92037','92109')
```

Pull census data about ZIP code areas

```{r}
reverse_zipcode(c('92037', "92109"))
```

## Pull data for all ZIP codes in the dataset

```{r}
zipdata <- reverse_zipcode(vax$zip_code_tabulation_area)
```

Take a look at the zipdata

```{r}
head(zipdata)
```

# Focus on the San Diego area

## Subset to San Diego county only areas

Install the "dplyr" package in R Console to work with postal ZIP codes by using the 'install.packages("dplyr")' command.

```{r, message=FALSE}
library(dplyr)
```

```{r}
sd <- filter(vax, county == "San Diego")

nrow(sd)
```

Take a look at the sd

```{r}
head(sd)
```
## Subset of all San Diego county areas with a population of over 10,000.

```{r}
sd.10 <- filter(vax, county == "San Diego" & 
                  age5_plus_population > 10000)
```

Take a look at the sd.10

```{r}
head(sd.10)
```

> **Q11. How many distinct zip codes are listed for San Diego County?**

```{r}
length(unique(sd$zip_code_tabulation_area))
```

> **A11: 107 distinct zip codes**

> **Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?**

```{r}
max_12_pop <- which.max(sd$age12_plus_population)
sd[max_12_pop, "zip_code_tabulation_area"]
```

> **A12: 92154**

**Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-11-15”?**

```{r}
date_2022.11.15 <- filter(vax, county == "San Diego" & 
                            as_of_date == "2022-11-15")

head(date_2022.11.15)
```

```{r}
overall_average <- function(x) {
x[is.na(x)] <- 0
mean(x)
}
```

```{r}
round(overall_average(date_2022.11.15$percent_of_population_fully_vaccinated)*100, 2)
mean(date_2022.11.15$percent_of_population_fully_vaccinated, na.rm = TRUE)
```

> **A13: 68.30%**

> **Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-11-15”?**

>**A14:**

```{r}
hist(date_2022.11.15$percent_of_population_fully_vaccinated, 
     main = "Histogram of Vaccination Rates Across San Diego County", 
     xlab = "Percent of Population Fully Vaccinated (as of 2022-11-15)", 
     ylab = "Count (Zip Code Areas)", 
     col = rainbow(10))
```

## Focus on UCSD/La Jolla

UC San Diego resides in the 92037 ZIP code area and is listed with an age 5+ population size of 36,144.

```{r}
ucsd <- filter(sd, zip_code_tabulation_area == "92037")
ucsd[1, ]$age5_plus_population
```


> **Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area**

>**A15:**

```{r}
library(ggplot2)
```

```{r, warning=FALSE}
ucsd_plot <- ggplot(ucsd) + 
  aes(x = ucsd$as_of_date, 
      y = ucsd$percent_of_population_fully_vaccinated) + 
  geom_point() + 
  geom_line(group = 1) +
  ylim(c(0,1)) +
  labs(title = "Vaccination Rate for La Jolla CA 92109", 
       x = "Date", y = "Percent Vaccinated")

print(ucsd_plot)
```

## Comparing to similar sized areas

Look across every zip code area with a population at least as large as that of the zip code 92037 (which has an age 5+ population size of 36,144) on as_of_date “2022-02-22”.

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 & 
                   as_of_date == "2022-11-15")

head(vax.36)
```

> **Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?**

> **A16:**

```{r}
mean_1 <- mean(vax.36$percent_of_population_fully_vaccinated)
mean_1
```

```{r, warning=FALSE}
ucsd_plot + 
  geom_hline(aes(yintercept = mean_1), color = "red", linetype = 2)
```

> **Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”?**

> **A17:**

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

>**Q18. Using ggplot generate a histogram of this data.**

> **A18:**

```{r, warning=FALSE}
ggplot(vax.36) + 
  aes(percent_of_population_fully_vaccinated) + 
  geom_histogram(bins = 30, color = "black", fill = rainbow(30)) + 
  xlim(0.00, 1.00) +
  labs(title = "Histogram of Vaccination Rate", 
       x = "Percent of Population Fully Vaccinated (as of 2022-11-15)", 
       y = "Count (Zip Code Areas with a Population as Large as La Jolla)")
```

> **Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?**

The average value:
```{r}
mean_1
```

For the 92109 zip code area:
```{r}
vax %>% filter(as_of_date == "2022-11-15") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```

For the 92040 zip code area:
```{r}
vax %>% filter(as_of_date == "2022-11-15") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```

> **A19: The 92109 and 92040zip code areas is below the average value.**

> **Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.**

> **A20:**

```{r, warning=FALSE}
vax.36.all <- filter(vax, age5_plus_population > 36144)


ggplot(vax.36.all) + 
  aes(as_of_date, 
      percent_of_population_fully_vaccinated, 
      group = zip_code_tabulation_area) + 
  geom_line(alpha = 0.2, color = "blue") + 
  ylim(0.00, 1.00) + 
  labs(x = "Date", 
       y = "Percent Vaccinated", 
       title = "Vaccination Rate Across California", 
       subtitle = "Only areas with a population ablove 36k are shown") + 
  geom_hline(yintercept = mean_1, linetype = 2)
```

> **Q21. How do you feel about traveling for Thanksgiving Break and meeting for in-person class afterwards?**

> **A21: I feel enthusiastic about the upcoming travel trip with my wife and my little girl during Thanksgiving Break, and I also feel excited about the in-person class afterward to learn new things.**

























